Modal Abstractions for Virtualizing Memory Addresses

### 2024-06-30 examples.tex
+

\begin{figure}\footnotesize
  \begin{lstlisting}[mathescape]
$\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi,m)  \ast \texttt{rbp-8} \mapsto_{\textsf{v}} \textsf{entry} \ast \texttt{r8}  \mapsto_{\textsf{r}} \textsf{\_} \ast \texttt{rdi}  \mapsto_{\textsf{r}} \textsf{\_} \ast \sumpv\paddr\qfrac\true  \sumapaces\rtv\delta  }_{\rtv}$
mov    -0x8[rbp],rdi
$\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi,m)  \ast \textsf{rbp-8} \mapsto_{\textsf{v}} \textsf{entry} \ast \texttt{r8}  \mapsto_{\textsf{r}} \textsf{\_} \ast \texttt{rdi}  \mapsto_{\textsf{r}} \textsf{entry} \ast \sumpv\paddr\qfrac\true  \sumapaces\rtv\delta   }_{\rtv}$
mov     rdi, r8
$\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi,m)  \ast \textsf{rbp-8} \mapsto_{\textsf{v}} \textsf{entry} \ast \texttt{r8}  \mapsto_{\textsf{r}} \textsf{entry} \ast \texttt{rdi}  \mapsto_{\textsf{r}} \textsf{entry} \ast \textsf{entry-KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{entry entry\_val} \ast \sumpv\paddr\qfrac\true  \sumapaces\rtv\delta  }_{\rtv}$
mov    [r8],rdi
$\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi,m)  \ast \textsf{rbp-8} \mapsto_{\textsf{v}} \textsf{entry} \ast \texttt{r8}  \mapsto_{\textsf{r}} \textsf{entry} \ast rdi  \mapsto_{\textsf{r}} \textsf{entry\_val} \ast \textsf{entry-KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{entry entry\_val} \ast \sumpv\paddr\qfrac\true  \sumapaces\rtv\delta  }_{\rtv}$
and    0x1,rdi
mov    rdi,rax
;;  if (!entry->present) {
cmp    0x0,rax
jne    161 <pte_get_next_table_succ+0xa1>
lea    -0x10[rbp],rdi
;;pte_initialize(&next);
callq  70 <pte_initialize>
...
;;}
\end{lstlisting}
\vspace{-1em}
\caption{Making sure that \textsf{entry}'s frame is referencing to an address backed up by a physical page.}
\label{fig:calltopteinitialize}
\end{figure}

As we see in Figure \ref{fig:loadentry}, we have the physical address (\textsf{entry}) of the page table entry kept in the register $r8$, but not the memory mapping to load the entry value referenced by the physical address \textsf{entry}. 
\subsubsection{Identity Mappings}
Most kernels maintain an invariant on their page tables that the virtual address of any page used for a page table lives at a virtual address
whose value is a constant offset from the physical address a practice sometimes referred to as \emph{identity mapping} (even though the physical-to-virtual translation
is typically not literally the identity function, but adding a constant offset).\footnote{Some kernels do this for all physical memory on the machine, simplifying interaction
with DMA devices.}
\begin{figure*}
\footnotesize
\[
\begin{array}{l}
  \mathcal{I}\textsf{ASpace}_{\textsf{id}}(\ptablestore,\Xi,m)\stackrel{\triangle}{=}\\ \textsf{ASpace\_Lookup}_{\textsf{id}}(\ptablestore,\Xi,m) \ast 
  \bigast{(\vaddr, \textsf{paddr})\in \ptablestore}{\exists\;(\textsf{l4e, l3e, l2e, l1e, paddr})\ldotp \textsf{L}_{4}\_\textsf{L}_{1}\_\textsf{PointsTo}(\vaddr\textsf{, l4e, l3e, l2e, l1e, paddr})} \\
  \bigast{(\paddr) \in \Xi}{\exists\; (\textsf{qfrac, q, val,}\vaddr) \ldotp \ulcorner \vaddr = \paddr + \textsf{OFF} \urcorner \ast  \underbrace{\sumwalkabs\vaddr\qfrac\paddr }_\text{Ghost translation} \ast 
% \qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad    
\underbrace{\paddr \mapsto_{\mathsf{p}}\{\textsf{qfrac}\}\; \vale}_\text{Physical location} \ast \ulcorner \textsf{qfrac} = 1 \leftrightarrow \; \lnot\textsf{entry\_present }(\vale) \urcorner }\\
  \textsf{ where } \\
   \textsf{ASpace\_Lookup}_{\textsf{id}}(\ptablestore,\Xi,m) \stackrel{\triangle}{=} \lambda\textsf{ cr3val} \ldotp \; \exists \gammaPred \; \ldotp \ulcorner m \; !!\; \textsf{cr3val} = \textsf{Some } \gammaPred \urcorner \ast
   \ownGhost\gammaPred{\authfull{\ptableabswalk\ptablestore}} \ast  \ownGhostpv\gammaPred{\authfull{\pvmapping\Xi}}

\end{array}
\]
\vspace{-1em}
\caption{Global Address-Space Invariant in Figure \ref{fig:peraspaceinvariant} extended with a ghost map bookkeeping identity mappings }
  \label{fig:peraspaceinvariant_with_p2v_extension}
  \end{figure*}
Now, to read the value referenced by physical address \textsf{entry} in Figure \ref{fig:loadentry}, first, we open our the addresss-space invariant ($\mathcal{I}\textsf{ASpace}_{\textsf{id}}(\ptablestore,\Xi,m)$), and obtain the virtual mapping out of the identity mapping $\Xi$ with the side condition

\begin{equation*}
 \ulcorner \textsf{qfrac} = 1 \leftrightarrow \; \lnot\textsf{entry\_present }(\vale) \urcorner \tag{*}
\end{equation*}

 assuring that looking up identity mappings are safe, because the full ownership asserts the non-presence of the entry which requires allocation a new physical page and setting the presence bit. By obtaining the identity mapping for \textsf{entry} in Figure \ref{fig:loadentry} -- ($\textsf{entry-KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{entry\_val}$ with page addr \textsf{entry}),
 \[
 \begin{array}{l}
\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi\setminus\{entry\},m) \ast \textsf{rbp-8} \mapsto_{\textsf{v}} \textsf{entry} \ast r8  \mapsto_{\textsf{r}} \textsf{entry} \textsf{\_} \ast rdi  \mapsto_{\textsf{r}} \textsf{entry} \ast \textsf{entry-KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{entry\_val} \ast \sumapaces\rtv\delta  \ast \sumpv\paddr\qfrac\true  }_{\rtv}
mov    [r8],rdi
\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi,m) \ast \textsf{rbp-8} \mapsto_{\textsf{v}} \textsf{entry} \ast r8  \mapsto_{\textsf{r}} \textsf{entry} \textsf{\_} \ast rdi  \mapsto_{\textsf{r}} \textsf{entry\_val} \ast  \textsf{entry-KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{entry\_val} \ast \sumapaces\rtv\delta  \ast \sumpv\paddr\qfrac\true  }_{\rtv}
\end{array}
 \]
 we are able to load the \textsf{entry\_val} into \textsf{rdi}, and check the presence bit via $and 0x1,rdi$.

>>>>

\begin{figure}\footnotesize
  \begin{lstlisting}[mathescape]
$\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi,m)  \ast \texttt{rbp-8} \mapsto_{\textsf{v}} \textsf{entry} \ast \texttt{r8}  \mapsto_{\textsf{r}} \textsf{\_} \ast \texttt{rdi}  \mapsto_{\textsf{r}} \textsf{\_} \ast \sumpv\paddr\qfrac\true  \sumapaces\rtv\delta  }_{\rtv}$
mov    -0x8[rbp],rdi
$\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi,m)  \ast \textsf{rbp-8} \mapsto_{\textsf{v}} \textsf{entry} \ast \texttt{r8}  \mapsto_{\textsf{r}} \textsf{\_} \ast \texttt{rdi}  \mapsto_{\textsf{r}} \textsf{entry} \ast \sumpv\paddr\qfrac\true  \sumapaces\rtv\delta   }_{\rtv}$
mov     rdi, r8
$\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi,m)  \ast \textsf{rbp-8} \mapsto_{\textsf{v}} \textsf{entry} \ast \texttt{r8}  \mapsto_{\textsf{r}} \textsf{entry} \ast \texttt{rdi}  \mapsto_{\textsf{r}} \textsf{entry}  \ast \sumpv\paddr\qfrac\true  \sumapaces\rtv\delta  }_{\rtv}$
$\specline{ \textsf{entry-KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{entry entry\_val} \ast \ulcorner  \textsf{qfrac} = 1 \leftrightarrow \; \lnot\textsf{entry\_present }\urcorner}$
mov    [r8],rdi
$\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi,m)  \ast \textsf{rbp-8} \mapsto_{\textsf{v}} \textsf{entry} \ast \texttt{r8}  \mapsto_{\textsf{r}} \textsf{entry} \ast rdi  \mapsto_{\textsf{r}} \textsf{entry\_val} \ast \sumpv\paddr\qfrac\true  \sumapaces\rtv\delta  }_{\rtv}$
$\specline{ \textsf{entry-KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{entry entry\_val} \ast \ulcorner  \textsf{qfrac} = 1 \leftrightarrow \; \lnot\textsf{entry\_present }  \urcorner }$
and    0x1,rdi
mov    rdi,rax
;;  if (!entry->present) {
cmp    0x0,rax
jne    161 <pte_get_next_table_succ+0xa1>
lea    -0x10[rbp],rdi
;;pte_initialize(&next);
callq  70 <pte_initialize>
...
\end{lstlisting}
\vspace{-1em}
\caption{Making sure that \textsf{entry}'s frame is referencing to an address backed up by a physical page.}
\label{fig:calltopteinitialize}
\end{figure}

As we see in Figure \ref{fig:calltopteinitialize}, we have the physical address (\textsf{entry}) of the page table entry kept in the register $\texttt{r8}$, but not the memory mapping to load the entry value referenced by the physical address \textsf{entry}. We need to construct this virtual mappings for \textsf{entry}.
\subsubsection{Identity Mappings}
Most kernels maintain an invariant on their page tables that the virtual address of any page used for a page table lives at a virtual address
whose value is a constant offset from the physical address a practice sometimes referred to as \emph{identity mapping} (even though the physical-to-virtual translation
is typically not literally the identity function, but adding a constant offset).\footnote{Some kernels do this for all physical memory on the machine, simplifying interaction
with DMA devices.}
\begin{figure*}
\footnotesize
\[
\begin{array}{l}
  \mathcal{I}\textsf{ASpace}_{\textsf{id}}(\ptablestore,\Xi,m)\stackrel{\triangle}{=}\\ \textsf{ASpace\_Lookup}_{\textsf{id}}(\ptablestore,\Xi,m) \ast 
  \bigast{(\vaddr, \textsf{paddr})\in \ptablestore}{\exists\;(\textsf{l4e, l3e, l2e, l1e, paddr})\ldotp \textsf{L}_{4}\_\textsf{L}_{1}\_\textsf{PointsTo}(\vaddr\textsf{, l4e, l3e, l2e, l1e, paddr})} \\
  \bigast{(\paddr) \in \Xi}{\exists\; (\textsf{qfrac, q, val,}\vaddr) \ldotp \ulcorner \vaddr = \paddr + \textsf{OFF} \urcorner \ast  \underbrace{\sumwalkabs\vaddr\qfrac\paddr }_\text{Ghost translation} \ast 
% \qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad    
\underbrace{\paddr \mapsto_{\mathsf{p}}\{\textsf{qfrac}\}\; \vale}_\text{Physical location} \ast \ulcorner \textsf{qfrac} = 1 \leftrightarrow \; \lnot\textsf{entry\_present }(\vale) \urcorner }\\
  \textsf{ where } \\
   \textsf{ASpace\_Lookup}_{\textsf{id}}(\ptablestore,\Xi,m) \stackrel{\triangle}{=} \lambda\textsf{ cr3val} \ldotp \; \exists \gammaPred \; \ldotp \ulcorner m \; !!\; \textsf{cr3val} = \textsf{Some } \gammaPred \urcorner \ast
   \ownGhost\gammaPred{\authfull{\ptableabswalk\ptablestore}} \ast  \ownGhostpv\gammaPred{\authfull{\pvmapping\Xi}}

\end{array}
\]
\vspace{-1em}
\caption{Global Address-Space Invariant in Figure \ref{fig:peraspaceinvariant} extended with a ghost map bookkeeping identity mappings }
  \label{fig:peraspaceinvariant_with_p2v_extension}
  \end{figure*}
Now, to read the value referenced by physical address \textsf{entry} in Figure \ref{fig:calltopteinitialize}, first, we open our the addresss-space invariant ($\mathcal{I}\textsf{ASpace}_{\textsf{id}}(\ptablestore,\Xi,m)$), and obtain the virtual mapping out of the identity mapping $\Xi$ with the side condition
\begin{equation*}
 \ulcorner \textsf{qfrac} = 1 \leftrightarrow \; \lnot\textsf{entry\_present }(\vale) \urcorner \tag{*}
\end{equation*}
 assuring that looking up identity mappings are safe, because the full ownership asserts the non-presence of the entry which requires allocation a new physical page and setting the presence bit. By obtaining the identity mapping for \textsf{entry} in Figure \ref{fig:calltopteinitialize} -- ($\textsf{entry-KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{entry entry\_val}$ with page addr \textsf{entry}),
 \[
 \begin{array}{l}
\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi\setminus\{entry\},m) \ast \textsf{rbp-8} \mapsto_{\textsf{v}} \textsf{entry} \ast r8  \mapsto_{\textsf{r}} \textsf{entry} \textsf{\_} \ast rdi  \mapsto_{\textsf{r}} \textsf{entry} \ast \textsf{entry-KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{entry\_val} \ast \sumapaces\rtv\delta  \ast \sumpv\paddr\qfrac\true  }_{\rtv}
mov    [r8],rdi
\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi,m) \ast \textsf{rbp-8} \mapsto_{\textsf{v}} \textsf{entry} \ast r8  \mapsto_{\textsf{r}} \textsf{entry} \textsf{\_} \ast rdi  \mapsto_{\textsf{r}} \textsf{entry\_val} \ast  \textsf{entry-KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{entry\_val} \ast \sumapaces\rtv\delta  \ast \sumpv\paddr\qfrac\true  }_{\rtv}
\end{array}
 \]
 we are able to load the \textsf{entry\_val} into \textsf{rdi}, and check the presence bit via $and 0x1,rdi$.
